---
title: "Analysis of taxonomic abundance and diversity in lung microbiome"
author: "MDP"
date: "2023-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of taxonomic abundance and diversity in lung microbiome

## Objectives

-   Separate datasets into samples with *Pseduomonas* spp. predominance or not
-   Define if predominance of *Pseduomonas* spp. in the microbiome of patients with Cystic Fibrosis(CF) and Non-Cystic fibrosis bronchiectasis (NCFB) affects the measures of diversity

## Preparation

### Command line: create summary files of taxonomy

The structure of the project repository is described below. All paths are relative to the `/script` folder where this script is located. The /processed_data folder contains all required data for downstream analysis.

```{bash Project dir structure, eval = FALSE}
.Rproject
├── notebook
├── processed_data
│   └── accessions
├── results
│   ├── tax_cf
│   └── tax_ncfb
└── scripts
```

After obtaining taxonomy classification using **nf-core** `taxprofiler_1.0.1` and `kraken2/bracken`, I use `kraken-biom_v1.2.0` to produce **biom** files necessary for analysis using `Phyloseq` in `R`.

```{bash Prepare .biom file, eval = FALSE}
  # starting at the root directory of repo
cd processed_data

  # taxpasta container is part of taxprofiler pipeline
for disease in {cf,ncfb}
    do 
    singularity exec $NXF_SINGULARITY_CACHE/kraken-biom_1.2.0.sif kraken-biom \
        --fmt json \
        --min G \
        -o ${disease}_kraken.biom \
        /scratch/mdprieto/results/cf_seed/taxprof_${disease}/bracken/k2_reports/* 
    done
```

### Import data into R

In case it is necessary, install the required packages.

```{r Packages install, eval = FALSE}
# ------- bioconductor
if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("phyloseq")

# ------- base R
req_packages <- c("ggplot2", "RColorBrewer", "patchwork","tidyverse")
  # check if already installed
not_installed <- req_packages[!(req_packages %in% installed.packages()[ , "Package"])]    
  # install only ones that are needed
if(length(not_installed)) install.packages(not_installed)    
```

Load all necessary libraries for analysis:

```{r Load packages, message = FALSE}
req_packages <- c("ggplot2", "RColorBrewer", "patchwork","tidyverse","phyloseq")
lapply(req_packages, require, character.only = TRUE)
```

Load **.biom** objects from `Bracken` taxonomic classification.

```{r Import .biom files}
# ------- import CF and NCFB biom files
cf_phyloseq <- import_biom("processed_data/cf_bracken.biom")
ncfb_phyloseq <- import_biom("processed_data/ncfb_bracken.biom")

# ------- adjust names to match tsv file
sample_names(cf_phyloseq) <- sub("_report", "", sample_names(cf_phyloseq))
sample_names(ncfb_phyloseq) <- sub("_report", "", sample_names(ncfb_phyloseq))

# ------- clean taxonomy table
  # CF phyloseq object
tax_table(cf_phyloseq) <- tax_table(cf_phyloseq) %>% sub("[a-z]__", "", .)
colnames(tax_table(cf_phyloseq)) <- c("kingdom", "phylum", "clade", "order", "family", "genus", "species")
  # NCFB phyloseq object
tax_table(ncfb_phyloseq) <- tax_table(ncfb_phyloseq) %>% sub("[a-z]__", "", .)
colnames(tax_table(ncfb_phyloseq)) <- c("kingdom", "phylum", "clade", "order", "family", "genus", "species")
```

Also import taxonomic classification summary, as it will be used to classify the samples where *Pseudomonas* spp. is the most abundant bacteria.

```{r Import taxonomic summaries}
# ------- import taxonomic summary excluding unnecessary columns
cf_summary_tsv <- read_tsv("results/tax_cf/CF_bracken_k2_db.tsv",
                   col_select = c(-3, -4))
ncfb_summary_tsv <- read_tsv("results/tax_ncfb/NCFB_bracken_k2_db.tsv",
                     col_select = c(-3, -4))

# ------- clean column names to reflect simple sample_id
colnames(cf_summary_tsv) <- sub("_SR.*", "", colnames(cf_summary_tsv))
colnames(ncfb_summary_tsv) <- sub("_SR.*", "", colnames(ncfb_summary_tsv))
```

## Predominance of *Pseudomonas* spp.

To make it easier to analyze the data, I pivot the table to a **long format** where every line has only one value assigned to an identifier. Furthermore, I calculate the relative abundance of a **taxa** in a new column.

Finally, I filter those cases where *Pseudomonas* spp. is the most abundant taxa in the microbiome and it represents more than 30% of the relative abundance.

```{r Predominance of Pseu in CF}
# ------- pivot to longer
  # do not pivot taxa_id columns and keep sample_ids together
cf_summary_tsv <- 
  cf_summary_tsv |>
  pivot_longer(cols = c(-1,-2),
               cols_vary = "slowest",
               names_to = "sample_id",
               values_to = "assigned_reads") 
  
# ------- calculate relative abundance of each species by sample_id
cf_summary_tsv <- 
  cf_summary_tsv |> 
  group_by(sample_id) |> 
  mutate(relative_abundance = assigned_reads / sum (assigned_reads))

# ------- define which samples have predominance of Pseudomonas
cf_pseudomonas <- 
  cf_summary_tsv |> 
  group_by(sample_id) |> 
  slice_max(n = 1,
            order_by = relative_abundance) |> 
  dplyr::filter(name == "Pseudomonas aeruginosa" & relative_abundance >= 0.3 ) |> 
  pull(sample_id)
```

The same process is performed for samples from NCFB patients

```{r Predominance of Pseu in NCFB}
# ------- pivot to long
ncfb_summary_tsv <- 
  ncfb_summary_tsv |> 
  pivot_longer(cols = c(-1,-2),
               cols_vary = "slowest",
               names_to = "sample_id",
               values_to = "assigned_reads") 
  
# ------- add relative abundances
ncfb_summary_tsv <- 
  ncfb_summary_tsv |> 
  group_by(sample_id) |> 
  mutate(relative_abundance = assigned_reads / sum (assigned_reads))

# ------- define predominance of Pseudomonas
ncfb_pseudomonas <- 
  ncfb_summary_tsv |> 
  group_by(sample_id) |> 
  slice_max(n = 1,
            order_by = relative_abundance) |> 
  dplyr::filter(name == "Pseudomonas aeruginosa" & relative_abundance >= 0.3 ) |> 
  pull(sample_id)
```

## Separate samples according to *Pseudomonas* predominance

We now use the vectors with samples that have predominance of _**Pseudomonas**_(pseu) 
in each of the data-sets to create two subsets for disease. After that, we remove
the old phyloseq objects to keep our workspace tidy.

```{r}
# ------- CF
cf_predominance_pseu <- subset_samples(cf_phyloseq, sample_names(cf_phyloseq) %in% cf_pseudomonas)  
cf_predominance_other <- subset_samples(cf_phyloseq, !sample_names(cf_phyloseq) %in% cf_pseudomonas)

# ------- NCFB
ncfb_predominance_pseu <- subset_samples(ncfb_phyloseq, sample_names(ncfb_phyloseq) %in% ncfb_pseudomonas)  
ncfb_predominance_other <- subset_samples(ncfb_phyloseq, !sample_names(ncfb_phyloseq) %in% ncfb_pseudomonas)

  # remove taxa summaries and subsetting vectors
rm(cf_summary_tsv, ncfb_summary_tsv, ncfb_pseudomonas, cf_pseudomonas)
```

## Calculate diversity measures according to predominance group

### CF

```{r}

```

