---
title: "Analysis of taxonomic abundance and diversity in lung microbiome"
author: "MDP"
date: "2023-08-16"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of taxonomic abundance and diversity in lung microbiome

## Objectives

-   Separate datasets into samples with *Pseduomonas* spp. predominance or not
-   Define if predominance of *Pseduomonas* spp. in the microbiome of patients with Cystic Fibrosis(CF) and Non-Cystic fibrosis bronchiectasis (NCFB) affects the measures of diversity

## Preparation

### Command line: create summary files of taxonomy

The structure of the project repository is described below. All paths are relative to the `/script` folder where this script is located. The /processed_data folder contains all required data for downstream analysis.

```{bash Project dir structure, eval = FALSE}

├── notebook
├── processed_data
│   └── accessions
├── results
│   ├── tax_cf
│   └── tax_ncfb
└── scripts
    └── .Rproject
    └── taxonomy_abundance.Rmd
```

After obtaining taxonomy classification using **nf-core** `taxprofiler_1.0.1` and `kraken2/bracken`, I use `kraken-biom_v1.2.0` to produce **biom** files necessary for analysis using `Phyloseq` in `R`.

### Prepare working environment

In case it is necessary, install the required packages.

```{r Packages install, eval = FALSE}
# ------- bioconductor

if (!require("BiocManager", quietly = TRUE))
    install.packages("BiocManager")

BiocManager::install("phyloseq")

# ------- base R

packages <- c("ggsignif", "tidyverse", "flextable", "vegan")

  # check if already installed
not_installed <- packages[!(packages %in% installed.packages()[ , "Package"])]

  # install only ones that are needed
if(length(not_installed)) install.packages(not_installed)    
```

Load all necessary libraries for analysis:

```{r Load packages, message = FALSE}
packages <- c("RColorBrewer", "phyloseq", "tidyverse", "flextable", "ggsignif")
lapply(packages, require, character.only = TRUE)
```

### Load phyloseq objects and standardize them

Load **.biom** objects from `Bracken` taxonomic classification. Creates two 
`phyloseq` objects, one for each disease.

```{r Import .biom files}
cf_ps <- import_biom("processed_data/cf_bracken.biom")
ncfb_ps <- import_biom("processed_data/ncfb_bracken.biom")
```

Now, we do a few steps to normalize the data by adding proper rank names, changing 
the sample names to match the ones used in taxonomic summary, normalizing 
the abundance according to sequencing depth, excluding taxa with low abundance 
(increases variance), and calculating the relative abundance of each taxa.

```{r}
# ------- adjust names to match tsv file

sample_names(cf_ps) <- sub("_report", "", sample_names(cf_ps))
sample_names(ncfb_ps) <- sub("_report", "", sample_names(ncfb_ps))

# ------- clean taxonomy table

  # clean taxa names in CF phyloseq object
tax_table(cf_ps) <- sub(pattern = "[a-z]__", replacement = "", 
                        x =tax_table(cf_ps) )
  # add names of taxonomic ranks 
colnames(tax_table(cf_ps)) <- c("Kingdom", "Phylum", "Clade",
                                "Order", "Family", "Genus", "Species")
  
  # repeat cleaning process for NCFB phyloseq object
tax_table(ncfb_ps) <- sub("[a-z]__", "", tax_table(ncfb_ps))
colnames(tax_table(ncfb_ps)) <- c("Kingdom", "Phylum", "Clade",
                                  "Order", "Family", "Genus", "Species")

# ------- adjust to sequencing depth

  # CF
total = median(sample_sums(cf_ps))
cf_ps = transform_sample_counts(cf_ps,
                                function(x, t=total) round(t * (x / sum(x))))
  # NCFB
total = median(sample_sums(ncfb_ps))
ncfb_ps = transform_sample_counts(ncfb_ps,
                                  function(x, t=total) round(t * (x / sum(x))))
  # clean environment variable
rm(total)

# # ------- remove reads associated with Human genome in Kraken2

cf_ps <- subset_taxa(cf_ps, !Genus=="Homo")
ncfb_ps <- subset_taxa(ncfb_ps, !Genus=="Homo")
```

### Import taxonomic summary table

Also import taxonomic classification summary, as it will be used to classify the samples where *Pseudomonas* spp. is the most abundant bacteria.

```{r Import taxonomic summaries}
# ------- import taxonomic summary excluding unnecessary columns
cf_summary_tsv <- read_tsv("../results/tax_cf/CF_bracken_k2_db.tsv",
                           col_select = c(-3, -4))
ncfb_summary_tsv <- read_tsv("../results/tax_ncfb/NCFB_bracken_k2_db.tsv",
                             col_select = c(-3, -4))

# ------- clean column names to reflect simple sample_id
colnames(cf_summary_tsv) <- sub(pattern = "_SR.*", replacement = "",
                               x = colnames(cf_summary_tsv))
colnames(ncfb_summary_tsv) <- sub("_SR.*", "", colnames(ncfb_summary_tsv))
```

### Categorize predominance of *Pseudomonas* spp.

To make it easier to analyze the data, I pivot the table to a **long format** where every line has only one value assigned to an identifier. Furthermore, I calculate the relative abundance of a **taxa** in a new column.

***Note:*** Finally, I filter those cases where *Pseudomonas* spp. is the most abundant taxa in the microbiome and it represents more than 30% of the relative abundance.

```{r Predominance of Pseu in CF}
# ------- pivot to longer
  # do not pivot taxa_id columns and keep sample_ids together
cf_predominance <- 
  cf_summary_tsv |>
  pivot_longer(cols = c(-1,-2),
               cols_vary = "slowest",
               names_to = "sample_id",
               values_to = "assigned_reads") 
  
# ------- define predominance of pseudomonas

cf_predominance  <- 
  cf_predominance |> 
  dplyr::group_by(sample_id) |> 
      # remove human assigned reads
  dplyr::filter(!grepl("omo sapiens", name)) |> 
    # calculate relative abundance after pruning human reads
  dplyr::mutate(relative_abundance = assigned_reads / sum (assigned_reads)) |> 
    # select top bacterial abundance
  dplyr::slice_max(n = 1, order_by = relative_abundance) |>
    # add factor to denote samples with predominance of pseudomonas
  dplyr::mutate(predominance = case_when(name == "Pseudomonas aeruginosa" & relative_abundance >= 0.3 ~ 'P. aeruginosa',
                                         TRUE ~ 'Other'))

  # remove summary_tsv to keep environment tidy
rm(cf_summary_tsv)
```

The same process is performed for samples from NCFB patients

```{r Predominance of Pseu in NCFB}
# ------- pivot to long

ncfb_predominance <- 
  ncfb_summary_tsv |> 
  pivot_longer(cols = c(-1,-2),
               cols_vary = "slowest",
               names_to = "sample_id",
               values_to = "assigned_reads") 
  
# ------- define pseudomonas predominance for NCFB

ncfb_predominance  <- 
  ncfb_predominance |> 
  dplyr::group_by(sample_id) |> 
  dplyr::filter(!grepl("omo sapiens", name)) |> 
  dplyr::mutate(relative_abundance = assigned_reads / sum (assigned_reads)) |>
  dplyr::slice_max(n = 1, order_by = relative_abundance) |>
  dplyr::mutate(predominance = case_when(name == "Pseudomonas aeruginosa" & relative_abundance >= 0.2 ~ 'P. aeruginosa',
                                         TRUE ~ 'Other')) 

  # remove summary_tsv to keep environment tidy
rm(ncfb_summary_tsv)
```

### Codify predominance in phyloseq objects

Here we add the grouping variable to our main phyloseq objects. Also, we filter out the low abundancy taxa (should not be done in dataset to use for alpha diversity) and produce a phyloseq object with relative abundances of different taxa

```{r Relative abundance and code predominance}
  # add predominance variable to sample_data
sample_data(cf_ps)$Predominance <- cf_predominance$predominance
sample_data(ncfb_ps)$Predominance <- ncfb_predominance$predominance


# ------- relative abundance
  
# eliminate taxa with poor representation and calculate relative abundance
  # NCFB
ncfb_rel_abundance <- 
  filter_taxa(ncfb_ps,
              function(x) sum(x > 3) > (0.2*length(x)), TRUE) |> 
  transform_sample_counts(function(OTU) OTU/sum(OTU))

  # CF
cf_rel_abundance  <-  
  filter_taxa(cf_ps,
              function(x) sum(x > 3) > (0.2*length(x)), TRUE) |> 
  transform_sample_counts(function(OTU) OTU/sum(OTU))
```

## Impact of _Pseudomonas aeruginosa_ predominance in diversity

### CF - Alpha diversity by predominance 

Produce a summary bar plot of all species according to predominance variable

```{r CF distribution of taxa}
cf_rel_abundance %>% 
    # summarize at phylum level
  plot_bar(x = "Sample",
           fill = "Phylum") +
    # scale = free_x eliminates empty columns
  facet_grid (~Predominance, scale = "free_x") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
    # adds italics to the title
  labs(title = expression("CF with"~italic(Pseudomonas)~"spp. predominance"))
```

I now calculate alpha diversity measurement and from the resulting plot obtain the summarized estimates too

```{r CF alpha diversity violin plot}

cf_alpha <- 
    # plots according to predominance
  plot_richness(cf_ps, x = "Predominance",
              measures = c("Chao1", "Shannon", "InvSimpson")) +
    # change to a violin plot
  geom_violin(aes(fill = Predominance)) +
    # manually define colors of violins
  scale_fill_manual(values = c("red3", "blue3")) +
  theme_light() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(), 
        strip.text.x = element_text(size = 14, colour = "black", face = "bold"),
        ) +
    # adds significance symbols of wilcoxon
  ggsignif::geom_signif(comparisons = list(c("P. aeruginosa", "Other")), 
                        map_signif_level = TRUE) ; cf_alpha
```


Then, using the results from `plot_richness()` which also summarize the richness value for all samples, I create a table with the mean **alpha diversity measurements** and the resulting Wilcoxon p_value

```{r CF alpha diversity table}

cf_alpha$data |>
    # group by alpha measure variable and predominance
  group_by(variable, Predominance) |>
  summarise(Mean = round(mean(value), 2)) |> 
    # manually add wilcoxon results 
  mutate("Wilcoxon (p_val)" = case_when(Predominance == "Other" ~ "< 0.001", 
                                        TRUE ~ NA)) |> 
  rename("Alpha measure" = "variable") |> 
    # format flextable to bold,centered headers and add colored rows
  flextable::flextable()  |> 
  flextable::theme_vanilla() |> 
  flextable::bg(i = ~ Predominance == "pseudomonas", bg = "#EFEFEF", part = "body") |> 
  flextable::align(align = "center", part = "header")
```

### NCFB - Alpha diversity by predominance 

The same analysis is repeated for the NCFB samples.

```{r NCFB alpha diversity plots}
=======
### NCFB

The same analysis is repeated for the NCFB samples.

```{r}
>>>>>>> ccda4328cacbd6d9d53439c92e694d025cc64770
# ------- bar_plot of taxa relative abundances

ncfb_rel_abundance %>% 
  plot_bar(x = "Sample",
           fill = "Phylum") +
  facet_grid (~Predominance, scale = "free_x") +
  theme(axis.text.x = element_blank(), axis.ticks.x = element_blank()) +
  labs(title = expression("NCFB with"~italic(Pseudomonas)~"spp. predominance"))

# ------- plot summary of alpha diversity measures

ncfb_alpha <- 
  plot_richness(cf_ps, x = "Predominance",
              measures = c("Chao1", "Shannon", "InvSimpson")) +
  geom_violin(aes(fill = Predominance)) +
  scale_fill_manual(values = c("red3", "blue3")) +
  theme_light() +
  theme(axis.text.x = element_blank(),
        axis.title.x = element_blank(), 
        strip.text.x = element_text(size = 14, colour = "black", face = "bold"),
        ) +
  ggsignif::geom_signif(comparisons = list(c("P. aeruginosa", "Other")), 
                        map_signif_level = TRUE); ncfb_alpha

# ------- table summary of alpha diversity
ncfb_alpha$data |>
  group_by(variable, Predominance) |>
  summarise(Mean = round(mean(value), 2)) |> 
  mutate("Wilcoxon (p_val)" = case_when(Predominance == "Other" ~ "< 0.001", 
                                        TRUE ~ NA)) |> 
  rename("Alpha measure" = "variable") |> 
  flextable::flextable()  |> 
  flextable::theme_vanilla() |> 
  flextable::bg(i = ~ Predominance == "pseudomonas", bg = "#EFEFEF", part = "body") |> 
  flextable::align(align = "center", part = "header")
```

### CF Beta diversity

For both diseases, I measure beta-diversity using primarily the bray-curtis index

```{r}

  # create empty list of length = number of indexes to use
plot_list <-  vector("list", length = 2)
names(plot_list) <- c("bray", "jsd")

for (i in c("bray", "jsd")){
  
    ## ------ Prepare distances and ordinate object
      # Calculate distance matrix
    iDist <- distance(cf_rel_abundance, method=i)
      # Calculate ordination
    iMDS  <- ordinate(cf_rel_abundance, "MDS", distance=iDist)
    
    ## ------ Make plot
      # Don't carry over previous plot (if error, p will be blank)
    p <- NULL
      # Create plot, store as temp variable, p
    p <- plot_ordination(cf_rel_abundance, iMDS, color="Predominance", shape = "Predominance")
      # Add title to each plot
    p <- p + ggtitle(paste("MDS using distance method ", i, " for CF samples", sep=""))
      # Save the graphic to file.
    plot_list[[i]] = p
}

```

I calculate the magnitude of the difference using PERMANOVA in the `vegan` package and procude a clean plot for sharing

```{r permanova CF}

# calculate PERMANOVA for bray-curtis
vegan::adonis2(distance(cf_rel_abundance, method="bray") ~ Predominance,
              data = data.frame(sample_data(cf_rel_abundance)))

# produce quality plot of bray-curtis
plot_list[[1]] +
  stat_ellipse(aes(group=Predominance)) +
  theme_light() +
  labs(caption = "PERMANOVA p<0.001") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 18) )
```

### NCFB beta diversity

I repeat the same process for the NCFB data.

```{r}

  # create empty list of length = number of indexes to use
plot_list <-  vector("list", length = 2)
names(plot_list) <- c("bray", "jsd")

for (i in c("bray", "jsd")){
  
    ## ------ Prepare distances and ordinate object
      # Calculate distance matrix
    iDist <- distance(ncfb_rel_abundance, method=i)
      # Calculate ordination
    iMDS  <- ordinate(ncfb_rel_abundance, "MDS", distance=iDist)
    
    ## ------ Make plot
      # Don't carry over previous plot (if error, p will be blank)
    p <- NULL
      # Create plot, store as temp variable, p
    p <- plot_ordination(ncfb_rel_abundance, iMDS, color="Predominance", shape = "Predominance")
      # Add title to each plot
    p <- p + ggtitle(paste("MDS using distance method ", i, " for NCFB samples", sep=""))
      # Save the graphic to file.
    plot_list[[i]] = p
}

```

I calculate the magnitude of the difference using PERMANOVA in the `vegan` package and procude a clean plot for sharing

```{r permanova NCFB}

# calculate PERMANOVA for bray-curtis
vegan::adonis2(distance(ncfb_rel_abundance, method="bray") ~ Predominance,
              data = data.frame(sample_data(ncfb_rel_abundance)))

# produce quality plot of bray-curtis
plot_list[[1]] +
  stat_ellipse(aes(group=Predominance)) +
  theme_light() +
  labs(caption = "PERMANOVA p<0.001") +
  theme(plot.title = element_text(face = "bold", hjust = 0.5, size = 18) )

```
